Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2012-09-19 documents/mixture_models.tex
small tweaks

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we can; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\sdss\ Catalog galaxy models, which only include exponential,
de~Vaucouleurs, and composite (mixture of the two) radial profiles.
In detail, in fact, the \sdss\ Catalog models are small modifications
of these (\citealt{lupton}); details below.  So our goal here is to
provide replacements for these models in order to improve the
performance of galaxy image modeling and analysis software.  Some of
the models we present have been used previously in tools for precise
photometry (Bundy \etal, forthcoming); they could also be used to
speed other image-fitting systems, like the very successful
\project{GALFIT} (\citealt{galfit}).

@@

We wish to make an approximation to the two-dimensional circular
exponential profile $Q_{\exp}(\cdot)$
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\equiv& \exp(-\alpha_{\exp}\,[|\spos| - 1])
\\
\alpha_{\exp} &\equiv& -1.67834699
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and
$\alpha_{\exp}$ ensures that the profile has unit half-light radius.
The position $\spos$ is dimensionless because it parameterizes the
unit-size and unit-flux dimensionless function.  We seek the best
(where ``best'' will be defined below) $M_g$-Gaussian mixture
approximation
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\approx& \sum_{m=1}^{M_{\exp}} a_{\exp,m}\,\normal(\spos\given\zero,\var_{\exp,m})
\\
\var_{\exp,m} &\equiv& \sigma^2_{\exp,m}\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\var_m$ in the mixture are scalar multiples of the identity matrix
$\identity$ because we are requiring this dimensionless function to be
precisely circular (so every component is itself circular and
concentric).  Similarly for the de~Vaucouleurs profile
\begin{eqnarray}\displaystyle
Q_{\dev}(\spos) &\equiv& \exp(-\alpha_{\dev}\,[|\spos|^{1/4} - 1])
\\
\alpha_{\dev} &\equiv& -7.66924944
\\
Q_{\dev}(\spos) &\approx& \sum_{m=1}^{M_{\dev}} a_{\dev,m}\,\normal(\spos\given\zero,\var_{\dev,m})
\\
\var_{\dev,m} &\equiv& \sigma^2_{\dev,m}\,\identity
\quad .
\end{eqnarray}
(The half-light radius parameters $\alpha_{\exp}$ and $\alpha_{\dev}$
are taken from \citealt{ciotti}.)  The challenge we meet below is to
determine the parameters
\begin{eqnarray}
\{a_{\exp,m},\sigma_{\exp,m}\}_{m=1}^{M_{\exp}} \mbox{~and~} \{a_{\dev,m},\sigma_{\dev,m}\}_{m=1}^{M_{\dev}}
\end{eqnarray}
to best approximate the
traditional galaxy profile functions, under some sensible definiton of
the word ``best'', as a function of the model complexity parameters
(numbers of components) $(M_{\exp}, M_{\dev})$.

The \sdss\ pipelines make use of modified profiles, which have been
truncated smoothly at large radius and (in the case of the
de~Vaucouleurs profile) ``softened'' at the center (\citealt{lupton}).
The \sdss\ form of the exponential profile is
\begin{eqnarray}\displaystyle
Q_{\lux}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha_{\lux}\,[|\spos| - 1]) & \mbox{for~}|\spos| < 3 \\
  \exp(-\alpha_{\lux}\,[|\spos| - 1])
  \,\left[1 - [|\spos| - 3]^2\right]^2 & \mbox{for~}3 < |\spos| < 4 \\
  0                                   & \mbox{for~}4 < |\spos|
\end{array}\right.
\\
\alpha_{\lux} &\equiv& -1.67835
\quad ,
\end{eqnarray}
and the \sdss\ form of the de~Vaucouleurs profile is
\begin{eqnarray}\displaystyle
Q_{\luv}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha_{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right]) & \mbox{for~}|\spos| < 7 \\
  \exp(-\alpha_{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right])
  \,\left[1 - [|\spos| - 7]^2\right]^2 & \mbox{for~}7 < |\spos| < 8 \\
  0                                   & \mbox{for~}8 < |\spos|
\end{array}\right.
\\
\alpha_{\luv} &\equiv& -7.66925
\quad .
\end{eqnarray}
(The half-light radius parameters $\alpha_{\exp}$ and
$\alpha_{\dev}$---and the softening and cutoff radius parameters---are
taken from the \sdss\ codebase.)  The \sdss\ radial profiles are much
better behaved than the standard forms; we recommend them strongly.
Once again, the challenge is to determine the best parameters
\begin{eqnarray}
\{a_{\lux,m},\sigma_{\lux,m}\}_{m=1}^{M_{\lux}} \mbox{~and~} \{a_{\luv,m},\sigma_{\luv,m}\}_{m=1}^{M_{\luv}}
\quad .
\end{eqnarray}

>>>>

We need good performance in two-dimensional image synthesis (for
fitting) because, with \thetractor\ (Lang \etal, forthcoming), we are
building a comprehensive model of all the imaging data we have; this
will have models of many millions of galaxies in imaging that contains
on the order of $10^{13}$ pixels.  We are basing our models on the
\sdss\ Catalog galaxy models, which only include exponential,
de~Vaucouleurs, and composite (mixture of the two) radial profiles.
In detail, in fact, the \sdss\ Catalog models are small modifications
of these (\citealt{lupton}); details below.  So our goal here is to
provide replacements for these models in order to improve the
performance of galaxy image modeling and analysis software.  Some of
the models we present have been used previously in tools for precise
photometry (Bundy \etal, forthcoming); they could also be used to
speed other image-fitting systems, like the very successful
\project{GALFIT} (\citealt{galfit}).

@@

We wish to make an approximation to the two-dimensional circular
exponential profile $Q_{\exp}(\cdot)$
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\equiv& \exp(-\alpha_{\exp}\,[|\spos| - 1])
\\
\alpha_{\exp} &\equiv& -1.67834699
\quad ,
\end{eqnarray}
where $\spos$ is a dimensionless focal-plane position, and
$\alpha_{\exp}$ ensures that the profile has unit half-light radius.
The position $\spos$ is dimensionless because it parameterizes the
unit-size dimensionless function.  We seek the best (where ``best''
will be defined below) $M_{\exp}$-Gaussian mixture approximation
\begin{eqnarray}\displaystyle
Q_{\exp}(\spos) &\approx& \sum_{m=1}^{M_{\exp}} a_{\exp,m}\,\normal(\spos\given\zero,\var_{\exp,m})
\\
\var_{\exp,m} &\equiv& \sigma^2_{\exp,m}\,\identity
\quad ,
\end{eqnarray}
where all of the means are exactly zero and all of the variances
$\var_{\exp,m}$ in the mixture are scalar multiples of the identity
matrix $\identity$ because we are requiring this dimensionless
function to be precisely circular (so every component is itself
circular and concentric).  Similarly for the de~Vaucouleurs profile
\begin{eqnarray}\displaystyle
Q_{\dev}(\spos) &\equiv& \exp(-\alpha_{\dev}\,[|\spos|^{1/4} - 1])
\\
\alpha_{\dev} &\equiv& -7.66924944
\\
Q_{\dev}(\spos) &\approx& \sum_{m=1}^{M_{\dev}} a_{\dev,m}\,\normal(\spos\given\zero,\var_{\dev,m})
\\
\var_{\dev,m} &\equiv& \sigma^2_{\dev,m}\,\identity
\quad .
\end{eqnarray}
(The half-light radius parameters $\alpha_{\exp}$ and $\alpha_{\dev}$
are taken from \citealt{ciotti}.)  The challenge we meet below is to
determine the parameters
\begin{eqnarray}
\{a_{\exp,m},\sigma_{\exp,m}\}_{m=1}^{M_{\exp}} \mbox{~and~} \{a_{\dev,m},\sigma_{\dev,m}\}_{m=1}^{M_{\dev}}
\end{eqnarray}
to best approximate the traditional galaxy profile functions, under
some sensible definiton of the word ``best'', as a function of the
model complexity parameters (numbers of components) $(M_{\exp},
M_{\dev})$.

...In addition to these there are general Sersi\'c profiles, of which
the exponential and de~Vaucouleurs profiles are special cases...

The \sdss\ pipelines make use of modified profiles, which have been
truncated smoothly at large radius and (in the case of the
de~Vaucouleurs profile) ``softened'' at the center (\citealt{lupton}).
The \sdss\ form of the exponential profile is
\begin{eqnarray}\displaystyle
Q_{\lux}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha_{\lux}\,[|\spos| - 1]) & \mbox{for~}|\spos| < 3 \\
  \exp(-\alpha_{\lux}\,[|\spos| - 1])
  \,\left[1 - [|\spos| - 3]^2\right]^2 & \mbox{for~}3 < |\spos| < 4 \\
  0                                   & \mbox{for~}4 < |\spos|
\end{array}\right.
\\
\alpha_{\lux} &\equiv& -1.67835
\quad ,
\end{eqnarray}
and the \sdss\ form of the de~Vaucouleurs profile is
\begin{eqnarray}\displaystyle
Q_{\luv}(\spos) &\equiv& \left\{\begin{array}{ll}
  \exp(-\alpha_{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right]) & \mbox{for~}|\spos| < 7 \\
  \exp(-\alpha_{\luv}\,\left[[|\spos|^2 + 0.0004]^{1/8} - 1\right])
  \,\left[1 - [|\spos| - 7]^2\right]^2 & \mbox{for~}7 < |\spos| < 8 \\
  0                                   & \mbox{for~}8 < |\spos|
\end{array}\right.
\\
\alpha_{\luv} &\equiv& -7.66925
\quad .
\end{eqnarray}
(The half-light radius parameters $\alpha_{\exp}$ and
$\alpha_{\dev}$---and the softening and cutoff radius parameters---are
taken from the \sdss\ codebase.)
